Waggle dance distributions quantify collective foraging in honey bee colonies

### 2022-06-28 analysis/manuscript/methods.Rmd
made some updates for submittion to sci-advances

As our estimated proportion of scouts is continuos on the interval $[0, 1]$ we used the R package plsRbeta [@Bertrand2013] to conduct the partial least squares analysis and performed a beta regression on the results using the R package betareg [@Cribari-Neto2010]. As the betareg package only works on the open interval $(0, 1)$ the data, $x$, was transformed using the following equation: $\frac{x(n-1)+0.5}{n}, as outlined in the betareg package documentation. After analysis the data was back transformed to the original values for the plots in Fig 5.

>>>>

As our estimated proportion of scouts is continuos on the interval $[0, 1]$ we used the R package plsRbeta [@Bertrand2013] to conduct the partial least squares analysis and performed a beta regression on the results using the R package betareg [@Cribari-Neto2010]. As the betareg package only works on the open interval $(0, 1)$ the data, $x$, was transformed using the following equation: $\frac{x(n-1)+0.5}{n}$, as outlined in the betareg package documentation. After analysis the data was back transformed to the original values for the plots in Fig 5.

### 2022-06-28 analysis/manuscript/paper.Rmd
made some updates for submittion to sci-advances

---
title: "Waggle dance distributions quantify collective foraging in honey bee colonies"
# author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
csl: nature.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

Honeybee foraging is an extraordinary collective behaviour that is directed through the waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops. Here, we present a mathematical framework to explain how the intrinsic process of the waggle dance can allow a colony to exploit different landscape types. From waggle dance observations, we infer the proportion of dances performed by bees that forage individually, and those that forage collectively. By applying this methodology to waggle dance data from twenty hives, we show our model closely fits real-world honeybee foraging patterns and we demonstrate that colonies vary their use of waggle dance information across different landscapes. Our methodology provides a tool to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

No communication system is more celebrated than the honeybee waggle dance, yet there remains aspects of the behaviour which have not yet been quantified.

Within honeybee colonies, a series of simple rules determine when and how much bees perform the celebrated waggle dance (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1994A]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022], without requiring any individuals to go out and visit multiple resources to make a directly informed choice. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006], functioning as a collective that makes decisions as a single entity.

Despite its intricacy and precision, a wealth of research now shows that foraging bees frequently do not use dance information to find forage sites [@Gruter2009]. Accordingly, in many situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002; @Dornhaus2004 @Dornhaus2006; @Gruter2013; @Price2019]. This plasticity in the use of collective behaviour is intriguing because it may provide clues as to the selection pressures that were critical in the evolution of the dance communication system.

Identifying the circumstances in which bees use the dance is likely to be a key step towards understanding when it is likely to afford fitness benefits at the colony level. Despite the amount of research interest that has focused on the proximate mechanisms of collective decision-making via dance communication, the ultimate evolutionary drivers of its unique evolution within *Apis* -unparalleled in even a single other social insect- remain obscure [@Price2015]. While it is likely that the spatiotemporal distribution of forage was key, some studies find the benefits of dance communication to be realized only in challenging environments where resources are clumped or ephemeral [@Sherman2002; @Dornhaus2004; @Dornhaus2006], while others find no effects of landscape heterogeneity [@Nurnberger2017], or that dancing comes at a cost in challenging environments [@Price2019] and is most beneficial when species richness is high [@Donaldson-Matasci2012].

Attempts to identify the circumstances in which dance communication is important have been challenging because they require constant monitoring of real-world colony weight [@Kirchner1998; @Sherman2002; @Dornhaus2006; @Gruter2013; @Nurnberger2017; @Price2019] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. To do this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see [@Nurnberger2017]), particularly given that multiple landscape variables may interact to determine the utility of dance communication.

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The duration of the waggle run indicates the distance to the resource and the angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../results/figures/manuscript_figures/fig1.png")
```

Here, we present a method to determine the extent of waggle dance use in colony foraging within specific landscapes by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. Our method examines the distribution of waggle run durations reported on the dance-floor and infers from its shape the relative contribution of dances for resources identified by individual search, "scouting" versus dances for resources found by following dances, “recruitment” [@Seeley1995]. Previously, the relative amount of scouting and recruitment trips have only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Our methods provide an efficient means of quantifying collective behaviour with no requirement to manipulate hive orientations, assay foraging efficiency or document individuals' previous experience. As such, they can identify when colonies adopt a collective foraging strategy- a critical step towards understanding when such a strategy outperforms individual-level behaviour.

@@

```{r, fig.cap="Figure 2. **Simulating honey bee foraging**. In our simulation model with scouting only (a), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (b) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (c) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions"}
knitr::include_graphics("../results/figures/manuscript_figures/fig2.png")
```

On the dance-floor, the number of waggle runs performed for a resource depends on its profitability [@Seeley1994]. Honeybees achieve this by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994; @Seeley1994A]. By combining this profitability bias with the distributions identified in our simulations, we can accurately describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 3, see Methods for details)

@@

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012; @Schurch2019; @Samuelson2021] to extract waggle run durations. Using this data, for each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004; @Clauset2009].

@@

In the urban environment, the first principal component explained ~73% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5c). This component correlates positively with parks, allotments and cemeteries (8% of land cover; Table 1) and negatively with railways, however this accounts for ~1% of land cover (Table 1). The dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), has a significant negative correlation with the first principal component for some sites removed over the jackknife sampling (Supplementary Fig. 2.), however, it is overall not a significant contributor to the variance in the loadings (Fig. 5c.). Parks, allotments and cemeteries are typically forage-rich for honeybees [@Baldock2019], yet our model indicates more scouting trips in these landscapes. However, note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components (Supplementary Fig. 3.).

@@

Our results within the agri-rural sites are in agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006; @Price2019]. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the proportion of scouting trips change with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5). We arrived at this conclusion through the analysis of waggle dance data, using a mathematical model to fit the waggle dance durations against. This method provides a time and labour efficient methodology to quantify collective behaviour. We analysed an existing data set and thus provide a proof of concept for how this new toolkit can be used to evaluate factors influencing waggle dance recruitment. The methods described here can also provide a framework to start exploring how collective decision making can help to buffer the effects of landscape change, and in this case how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics such as in the highly urbanised conurbations.

@@

>>>>

---
title: "Waggle dance distributions quantify collective foraging in honey bee colonies"
# author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: paper-refs.bib
csl: science.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(showtext)
showtext_auto()

@@

Honeybee foraging is an extraordinary collective behaviour that is directed by the waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops. Here, we present a mathematical framework to quantify how use of the waggle dance shifts in different landscape types. By decoding dances inside hives, we infer the proportion that are performed by bees that forage individually, and those that forage collectively. By applying this methodology to data from twenty hives, we show our model closely fits real-world honeybee foraging patterns and we demonstrate that colonies vary their use of waggle dance information across different landscapes. Our methodology provides a tool to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

The extraordinary waggle dance communication system of honeybees (Apis spp.) is a key example of how social behaviour can take on a form extending beyond that of the individual units [@Sumpter2006], such that groups function as collectives that make decisions as a single entity. Within colonies, a series of simple rules determine when and how much bees perform the waggle dance (Fig. 1) mean that choices between feeding sites occur at the level of the group rather than the individual [@Seeley1994]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022]. This is without requiring any individuals to visit multiple resources to make a direct comparison. The resulting colony-level decisions are a prime example of how self-organized collective behaviour could increase the foraging efficiency of groups.

A wealth of research now shows that foraging bees frequently do not use dance information to find forage sites [@Gruter2009]. Accordingly, in many situations, colonies that are prevented from communicating via dances achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019]. This plasticity in the use of collective behaviour is intriguing because it may provide clues as to the selection pressures that were critical in the evolution of the dance communication system.

Identifying the circumstances in which bees use the dance is likely to be a key step towards understanding when it is likely to afford fitness benefits at the colony level. Despite the amount of research interest that has focused on the proximate mechanisms of collective decision-making via dance communication, the ultimate evolutionary drivers of its unique evolution within Apis -unparalleled in even a single other social insect- remain obscure [@Price2015]. While it is likely that the spatiotemporal distribution of forage was key, some studies find the benefits of dance communication to be realized only in challenging environments where resources are clumped or ephemeral [@Sherman2002;@Dornhaus2004;@Dornhaus2006], while others find no effects of landscape heterogeneity [@Nurnberger2017]. Others have found that dancing comes at a cost in challenging environments [@Price2019] and is most beneficial when species richness is high [@Donaldson-Matasci2012]. However, identifying how variation in landscape structure is associated with waggle dance use is practically difficult.

Attempts to identify the circumstances in which dance communication is important have been challenging because they require constant monitoring of real-world colony weight [@Sherman2002;@Dornhaus2004;Gruter20013;@Price2019;@Dornhaus2006;@Nurnberger2017;@Kirchner1998] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. To do this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see [@Nurnberger2017]), particularly given that multiple landscape variables may interact to determine the utility of dance communication.

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The duration of the waggle run indicates the distance to the resource and the angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993;@Boch1956;@Esch1961;@Seeley1994;@Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../results/figures/manuscript_figures/fig1.png")
```

Here, we present a method to determine the extent of waggle dance use in colony foraging within specific landscapes by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. Our method examines the distribution of waggle run durations reported on the dance-floor and infers from its shape the relative contribution of dances for resources identified by individual search, “scouting” versus dances for resources found by following dances, “recruitment” [@Seeley1995]. Previously, the relative amount of scouting and recruitment trips have only been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Our methods provide an efficient means of quantifying collective behaviour with no requirement to manipulate hive orientations, assay foraging efficiency or document individuals' previous experience. As such, they can identify when colonies adopt a collective foraging strategy- a critical step towards understanding when such a strategy outperforms individual-level behaviour.

@@

```{r, fig.cap="Figure 2. **Simulating honey bee foraging**. In our simulation model with scouting only (a), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (b) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (c) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions. The scout distribution is best fit by an exponential (grey fit line), the recruit distribution by a Rayleigh distribution (black line)."}
knitr::include_graphics("../results/figures/manuscript_figures/fig2.png")
```

On the dance-floor, the number of waggle runs performed for a resource depends on its profitability [@Seeley1994]. Honeybees achieve this by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994;@Seeley1994A]. By combining this profitability bias with the distributions identified in our simulations, we can accurately describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 3, see Methods for details)

@@

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed a pre-existing dataset of 2827 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives that contributed to this dataset had been situated at different locations in South East England (see Methods, figure 4A) and visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually [@Couvillon2012;@Schurch2019;@Samuelson2021] to extract waggle run durations. Using this data, for each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter $p$. In each case, We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004;@Clauset2009].

@@

In the urban environment, our findings were more equivocal. The first principal component explained ~73% of the variance in the proportion of scouts (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5c). This correlated negatively with the dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), as would be expected if the proportion of scouts decreases when rich forage is available, since such land includes a high proportion of gardens. However, this relationship is very sensitive to site removal through the jackknife sampling (Supplementary Fig. 2.), and is not a significant contributor to the variance in the loadings (Fig. 5c.). Furthermore, the first principal component also correlates positively with parks, allotments and cemeteries (8% of land cover; also railways but these constitute ~1% of land cover; Table 1), which would typically be considered forage-rich for honeybees [@Baldock2019] (although note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components; Supplementary Fig. 3.).

@@

Our results within the agri-rural sites are in agreement with other studies which evaluated foraging performance with and without the dance [@Dornhaus2006;@Price2019]. These results suggest that, even though collective foraging is not always beneficial, in environments where high quality resources are present, but relatively scarce, collective foraging is being carried out by colonies, suggesting that under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the proportion of scouting trips change with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5). We arrived at this conclusion through the analysis of waggle dance data, using a mathematical model to fit the waggle dance durations against. This method provides a time and labour efficient methodology to quantify collective behaviour. We analysed an existing data set and thus provide a proof of concept for how this new toolkit can be used to evaluate factors influencing waggle dance recruitment. The methods described here can also provide a framework to start exploring how collective decision making can help to buffer the effects of landscape change, and in this case how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics such as in the highly urbanised conurbations.

@@

# Materials and Methods

## Data collection

Methods for how the data used for this study was collected can be found in full in Materials and Methods sections 2.1, 2.2 and 2.3 of [@Samuelson2021].

### Land-use preference analysis

Methods for how land-use types were classified can be found in Materials and Methods sections 2.6 of [@Samuelson2021].

### Waggle dance decoding

Methods for waggle dance decoding are fully described in Materials and Methods section 2.4 in [@Samuelson2021].

## Simulation

All simulation code was written in Python version 3.9 and uses the Pandas [@Mckinney2011] and Scipy [@Jones2001] packages.

A circular environment is first created with radius $r = 2.5$. The number of resources in the environment is generated as a random Poisson variable with rate equal to 5000 multiplied by the area of the environment. These are placed on polar coordinates with a uniformly selected angle, $\theta$, between 0 and 2$\pi$ and a radial value, $\rho$, between 0 and $r$, determined from the square root of the uniform position values multiplied by $r$. These polar coordinates are converted to Cartesian coordinates. Each location is then assigned to an instance of a resource object along with a random quality of between 0 and 10. This quality is combined with the distance of the resource to the centrally located hive to form a measure of how profitable the resource is (see model, equation?).

One-hundred honeybee objects are created, 20 of which 20 are on scouting trips and the rest recruited to follow scout dances. Scouts leave the environment following a random path through the environment generated by sampling a uniform random step length and angle. The number of paths the scout draws when searching is also determined as a uniform random number. Each straight line in the random path is converted to a rectangle with length equal to the path section length and a constant width of ~0.01 to represent an area the scout searches along that path. Of all the resources contained in the boxes drawn from the scout's path, the one closest to the colony is selected as the resource patch that the scout will report and will communicate its location if the quality of the resource resource exceeds a minimum threshold. Communication is simulated by pooling together all the resource patches found. If no resources are contained in the scout's path, they will not add any resources to the scout pool and draw a new path in the next foraging iteration.

Recruits represent honeybee objects which do not perform the searches the scouts do. Instead, they sample from the pool of resource objects reported by scouts. This sampling is done by selecting resources with a probability which is skewed towards the profitability of the resource, meaning more profitable resources have a greater chance of being selected by recruits. Recruits will then visit these resources and in the next iteration will add their resource to the pool of scout dances. Consequently, the pool of dances represents resources discovered by scouts and resources exploited by recruits. When a resource is depleted, it is removed from the environment and so any foragers that were foraging on it would select a different resource from the dance floor.

The simulation was run 100 times and every 5 time steps all distances reported by scouts and recruits were recorded and combined. This was done to reflect the way foraging data is collected in real honeybee studies. We fit an exponential and minimum of an exponential distribution to both the distribution of foraging distances reported by the scout and recruit objects. Fitting was done by deriving the maximum likelihood estimate for each model fit on each data source through their analytical solutions: $\hat{\lambda} = \frac{1}{\bar{x}}$, minimum of the exponential with a minimum foraging distance: $\hat{\lambda} = \frac{1}{\pi \bar{x^{2}}}$. As the exponential assumes distributions start from 0 the data was transformed to start from 0 by subtracting the minimum foraging distance from all foraging distances ($x = x - min(x)$) before fitting.

## Model

To describe the distribution of dance durations on the dance floor we formulated a generic model the duration of waggle dances. In the model resource patches are assumed to be randomly placed in the environment. Foragers scout for these patches. The rationale of the model is illustrated in Fig. 3: upon visiting a resource patch, foragers translate the profitability of a resource into the number of repeats of the dance. The number of repeats of the dance is a function of quality an distance. Recruits sample random dances and report the location of successful visits to resource patches on the dance floor. Through the feedback and over-representation of profitable resources on the dance floor recruits will converge to visiting the most profitable resource in vicinity of the hive. The distribution of dance durations is the superposition of scouting and recruiting trips.

As the resources patches are randomly placed in the environment, the distance after which the first resource is discovered approximately follows an exponential distribution (given by $\lambda e^{- \lambda x}$). Through the feedback mechanism that the dance floor provides, the colony can, collectively, locate the most profitable resource in its environment. For randomly placed resources in a two dimensional environment the distance to the nearest point is distributed according to a Rayleigh distribution (given by $2 \lambda \pi x e^{- \pi \lambda x^2}$)) [@Pyke1978]. Our simulation model shows that this describes the distances at which recruits visit resources well.  Knowing the distance distributions of scout and recruit trips we then assume that the a proportion $p$ of all trips are scout trips. With this information we can specify the distributions of distances on the dance floor (see Supplementary Material for details).

We implemented this in a full model which describes the distance distribution of the environment has $n$ different resource types (See Supplementary Material). However, in the full model the number of parameters increases with $2n$.  Even if the number of resources is low, it turned out to be cumbersome to estimate the parameters and the model tends to overfit. To facilitate estimation of the parameter $p$ we therefore used a simplified model to estimate the fraction of scout trips, where $m$ represents the lowest duration considered, and here a minimum waggle run duration in the data set.

For the simplified the model we assumed that the number of dances depends weakly on distance and there is a sizable quality differences between resources of a non-negligible size and that there is a sizable intensity of the high quality resource (See Supplementary Material for detailed derivation). Foragers on scouting trips are more likely to report larger distances than foragers on recruiting trips. By linearising the function that translates the profitability into the number of waggle dance run for the largest dance duration and normalising, we arrive at simplified distribution for dance durations for scouting trips:

$$ f_s(x)=a_s \left[ M_s^{-1} b_s e^{-b_s a_s (x-m)}  \left[ 1-a_s x \right]_+ \right] $$

where we used the shorthand ${x}_+=\max(x,0)$ The parameter $a_s^{-1}$ is the maximum dance duration by scouts, $a_s b_s$ is the intensity of resources found by scouts and $M_s=(1-a_s m)-b_s^{-1} \left(1-e^{-b_s(1-a_s m)}\right)$ is factor that normalises the distribution.

Recruit trips will be predominantly to high quality resources. Only if the nearest high quality patch is very far away will there be a more profitable patch of lesser quality available, and this happens only rarely if the intensity of the best quality resource is sizable. After linearising the function that translates the profitability into the number of waggle dance runs for short dance durations and normalising the distribution of dance durations reported from recruit trips in the simplified model is:

$$f_r(x)=M_r^{-1} 2 \pi a_r^2   b_r x e^{- \pi  b_r  (a_r x)^2} \left[1- a_r x\right]_+$$

where

$$M_r={(1-a_r m)e^{-\pi br(a_r m)^2} + \frac{erf(a_r \sqrt{\pi b_r m}) - erf(\sqrt{\pi b_r}))}{2 \sqrt{b_r}}}$$

is the normalisation factor, the parameter $a_r$ is the rate with which dances repeats depends on distance for recruit trips and $a_r^2 b_r$ is the intensity of high quality resources reported by recruited foragers.

The simplified distribution function is

$$P(\underline x= x )=p f_s(x)+ (1-p) f_r(x), $$ which we used for parameter estimation and model fittings.

## Statistical analysis

All analysis code is written in R [@RCore2020].

### Model fitting

All models are fit using Maximum likelihood estimation [@Burnham2002] by summation of the log of the simplified distribution function outlined in the methods section: model. The numerical optimisation routine is written in c++ and uses the Nelder-Mead simplex algorithm [@Nelder1965] implemented in the 'NLopt' library [@Johnson2020] and interfaced to R [@RCore2020] using 'Rcpp' [@Eddelbuettel2011].

The most parsimonious model is assessed using Akaike information criterion (AIC) [@Burnham2002; @Aho2014] and Akaike weights. The model with the lowest AIC score is deemed to be the most parsimonious.

Goodness of fit is assessed using the two-sample Kolmorgorov-Smirnov (KS) test [@Goldstein2004] and implemented in R using the ks.boot function of the package 'Matching' in R [@Sekhon2011].

### Partial Least Squares analysis

Prior to conducing the PLS we removed any sites in which the models fit was significantly different to the waggle dance durations for that site. This resulted in one agri-rural site (ROT) and no urban sites being removed from the analysis.

As our estimated proportion of scouts is continuos on the interval $[0, 1]$ we used the R package plsRbeta [@Bertrand2013] to conduct the partial least squares analysis and performed a beta regression on the results using the R package betareg [@Cribari-Neto2010]. As the betareg package only works on the open interval $(0, 1)$ the data, $x$, was transformed using the following equation: $\frac{x(n-1)+0.5}{n}$, as outlined in the betareg package documentation. After analysis the data was back transformed to the original values for the plots in Fig 5.

For the jackknifed resampling we iterated through the each site and removed it from the pool of data and then ran the PLS as described above, recoding the loadings for each iteration (see Supplementary Material for loadings with each site removed). The PLS loadings for each land-use type are plotted as a box plot in Fig 5. to show the spread of these variable types. A loading was determined to be significantly correlating with the first principal component if contributed more than its expected variance.

# Acknowledgments

**Funding**

This work was supported by the Biotechnology and Biological Sciences Research Council (BBSRC) through grant BB/M011178/1.

**Author Contributions**

Conceptualization: JP, RG, EL, VJ.
Methodology: JP, EL, VJ.
Software: JP.
Formal analysis: JP.
Investigation: AS, EL.
Writing - Original Draft: JP, VJ.
Writing - review & editing: JP, RG, EL, VJ.
Supervision: RG, EL, VJ.

**Competing interests**

Authors declare they have no competing interests.

**Data and materials availability**

All data and code is available on zenodo at....
